---
title: "EML-basic-test"
author: "Ivan C Hanigan"
date: "13/07/2014"
output: html_document
---

Writing R data into EML
-----------------------


`EML` now extends R's `data.frame` class by introducing the `data.set`
class which includes additional metadata required by EML.  A `data.set`
can be created much like a `data.frame` by specifying additional arguments

```{r}
require(EML)
```

```{r}
dat = data.set(river = c("SAC",  "SAC",   "AM"),
               spp   = c("king",  "king", "ccho"),
               stg   = c("smolt", "parr", "smolt"),
               ct    = c(293L,    410L,    210L),
               col.defs = c("River site used for collection",
                            "Species common name",
                            "Life Stage",
                            "count of live fish in traps"),
               unit.defs = list(c(SAC = "The Sacramento River",
                                  AM = "The American River"),
                                c(king = "King Salmon",
                                  ccho = "Coho Salmon"),
                                c(parr = "third life stage",
                                  smolt = "fourth life stage"),
                                "number"))

```


- `col.defs`: These are usually just plain text definitions, though a
  URI to a semantic definition can be particularly powerful. See "Advanced
  Use" for details on adding richer information, such as the method used to
  collect the data or set the geographic, taxanomic, or temporal coverage
  of an individual column.

- `unit.defs`:   For factors, this is a definition of
  the levels involved.  For numeric data, specify the units from [this
  list](http://knb.ecoinformatics.org/software/eml/eml-2.1.1/eml-unitTypeDefinitions.html#StandardUnitDictionary).
  For dates, specify the format, (e.g. YYYY or MM-DD-YY). For character
  strings, a definition of the kind of string can be given, (e.g. species
  scientific name), otherwise the column description will be used.

Alternatively, annotations can be added to an existing data frame,
`data.set(my.data.frame, col.defs = ..., unit.defs = ...)`.


We will also specify a default creator who will also be used as the
contact person for EML files created in this session.


```{r}
eml_config(creator="Carl Boettiger <cboettig@gmail.com>")
```


While we will always be able to provide alternative or additional creators
or contact person later, `eml_config` can store the defaults to save
us the trouble of entering this data repeatedly.  See `?eml_config`
for more information and other configurable defaults.


With this information in place, we have all the required metadata to
generate a minimally valid EML file documenting this dataset.



```{r}
eml_write(dat, file = "EML_example.xml")
```

```
[1] "EML_example.xml"
```


*for convenience, had we ommitted any essential metadata, such as
providing only an unannotated `data.frame` in place of a `data.set`,
`EML` will launch its metadata wizard to assist in constructing the
metadata based on the data.frame provided*.  While this may be helpful
starting out, regular users will find it faster to define the columns
and units directly in the format above.

See the [EML
generated](https://github.com/ropensci/EML/tree/master/inst/doc/EML_example.xml)
by this example.


We can validate the resulting document to check that no required
fields are missing and that no entry has been incorrectly formatted.
This ensures that other software can correctly parse and interpret the
metadata document:


```{r}
eml_validate("EML_example.xml") 
```
